Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-08-23 latex/main.tex
Whitespace
Fixup some minor formatting issues

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available 
(\ref{fig:methodology}; for details, see STAR Methods). 
First, sc2ts infers likely “copying paths” connecting each sample in the batch to node(s) in the current ARG 
(allowing recombination) using the Li and Stephens (LS) model \cite{Li2003}, 
which is a Hidden Markov Model (HMM) widely used in human genomics 
to approximate the effects of mutation and recombination on genetic inheritance. 
As recombination is relatively rare in viruses, 
we would expect most samples to copy from a single parent in the ARG. 
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path 
between the parents resulting in different genomic segments inherited from the different parents. 
Next, for clusters within the batch that are inferred to copy from the same ARG node, 
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster. 
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and 
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

@@

To illustrate the power of these libraries, we provide some examples 
in which we compare the Usher and sc2ts intersections in tskit format described earlier. 
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments. 
The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences 
represents 31.34 GB of nucleotide calls. 
Of the 80,574,518 missing data calls (Ns) in the alignments, 
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%). 
Additionally, 413,303 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code). 
UShER imputed only 4 calls from this set incorrectly, 
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations. 
This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees, 
required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 GB 
(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format). 
Performing this analysis using matUtils would require exporting the full dataset to VCF, 
and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments. 
In contrast with matUtils and most phylogenetic software, 
which provide a command line interface and suite of predefined operations that typically output text files, 
tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}. 
As an illustration of the open-ended and efficient nature of the analyses this enables, 
we computed the number of polytomies (nodes that have > 2 children) 
and the maximum number of children per node in the Usher and sc2ts outputs. 
In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148) 
and compute the maximum number of children (5918). 
In the sc2ts ARG, it took 385 ms and 6 lines of code to find the maximum number of polytomies (137,001) and 
maximum number of children (7708) across all 348 local trees.

@@

Recombination is an important evolutionary force in viruses \cite{Pérez-Losada2015}, 
but has been understudied in part due to a lack of suitable data models, methods, and tools \cite{Neches2020}. 
While the methods and tools developed here are tailored for SARS-CoV-2, 
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is 
that they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops. 
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem, 
making them highly adaptable for a broad range of applications. 
Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers, 
enabling and accelerating studies of SARS-CoV-2 for years to come.

@@

\section*{MAIN FIGURE TITLES AND LEGENDS}

%%%  At final submission, figure files MUST be 
%%%  provided separately as high-resolution image 
%%%  files. All of the panels for a figure should 
%%%  be in the same file. Figures should have 
%%%  clear labels/file names (Figure 1, Figure 2, 
%%%  etc.).

%%%  Figure titles and legends should be placed 
%%%  at the end of the main text. You do not 
%%%  need to place the figures, nor their titles 
%%%  and legends, within the main text. While 
%%%  typesetting your article, our team will 
%%%  place each figure in the best location 
%%%  based on the final layout and on your 
%%%  figure citations, e.g., (Figures 1A and 1B).

%%%  Please review the figure guidelines before 
%%%  submitting your final materials: 
%%%  https://www.cell.com/figureguidelines.

\noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}

\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}

The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
\newline
(A) Or each panel or group of panels can be described separately, as shown here and below.
\newline
(B) Graph of X, Y, and Z.
\newline
(C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
\newline
Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}. Example figure created by Cassie Comeau, Cell Press.

\section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}

%%%  Whenever possible, we encourage you to submit 
%%%  your main-text tables as Microsoft Word documents, 
%%%  using Word's table function. This will ensure 
%%%  the best results during conversion. Tables 
%%%  should be numbered Table 1, Table 2, etc. and 
%%%  should not include subpanels (do not use Table 1A, 
%%%  1B, etc.). Give each table a brief descriptive 
%%%  title. Table legends are optional but encouraged. 
%%%  Footnotes (superscript lowercase letters) should 
%%%  be used where necessary to indicate some feature 
%%%  of the data; please do not use bold, italic, 
%%%  colored text, or shading for this purpose. Use 
%%%  separate cells, not line breaks or spaces, for 
%%%  all discrete data elements. Small embedded 
%%%  graphics with color are OK.

%%%  Like figures, all tables must be cited within 
%%%  the main text, and our typesetting team will 
%%%  place the tables within the typeset paper at 
%%%  the appropriate locations.

\subsection*{Table 1. A table with clear organization of data}

\begin{tabular}{|l | l | l | l|} 
 \hline
 \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex] 
 \hline
 Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex] 
 \hline
 Row B & 7 & 78 & 5,415 \\ [1ex] 
 \hline
 Row C & 545 & 778 & 7507 \\ [1ex] 
 \hline
 Row D & 545 & 18,744 & 7,560 \\ [1ex] 
 \hline
 Row E & 88 & 788 & 6,344 \\ [1ex] 
 \hline
\end{tabular}

The table legend (optional) follows the table itself. 
The legend should be used to provide additional info that relates to the table as a whole.

\textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, 
such as this footnote to the first row (row A). Do not use footnotes in the table title.

\textsuperscript{b}More footnotes

@@

\subsubsection{Li and Stephens haplotype-copying model}

@@

In our early attempts to build ARGs, we explored using several values of $k$. 
We ran sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020 \cite{Zhan2023}}(REF). 
We would expect that a good value for $k$ should lead to parsimonious ARGs 
that capture plausible recombination events without incurring too many mutations, particularly reversions. 
A value of $k$ too low could result in many artifactual recombination events, 
whereas a value of $k$ too high could miss many genuine recombination events. 
We found that ARGs built with $k$ set to 3 contained many dubious recombination events, 
but ARGs built with $k$ set to 5 failed 
to include the well supported recombination events documented in a previous study \cite{Jackson2021}. 
Hence, we decided to use $k$ set to 4 to infer the final ARG described in this study.

@@

Retrospective sample grouping successfully captured most saltation lineages, 
but integrating Delta and Omicron into the ARG required further manual intervention. 
To ensure that Delta and Omicron are added, we introduce a third mechanism, “unconditional seeding”. 
The idea is to choose suitable samples belonging to each VOC, 
and attach them to the ARG unconditionally (i.e. without filtering using HMM cost). 
We sought “seed” samples for Delta, Omicron BA.1, and BA.2. 
For Delta, we used two sets of seeds: 
(1) 8 samples assigned to B.1.617.2, and 
(2) 9 additional samples assigned to its sibling lineage B.1.617.1. 
Specifically, these samples are the earliest samples present in the Viridian v04 dataset 
that were used for the designation of the VOCs. 
For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively. 
We chose SRR17041376 (a sample dated 2021-11-12 from South Africa) 
as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264), 
as it could be well explained as as a non-recombinant given the ARG available on its collection date. 
For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa), 
which is the earliest BA.2 sample present in the Viridian v04 dataset.
It is dated later than SRR17041376 (the BA.1 seed), 
allowing time for the BA.1 lineage to accumulate enough sampled descendants 
and to get established in the ARG. 
For the full list of seeds, 
see https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml.

@@

Sites which are enriched for systematic sequencing errors or frequently experience homoplasy 
can cause artifactual signals of recombination (REF). 
The bioinformatics community has identified and curated a set of problematic sites 
that are recommended to be excluded before phylogenetic inference (REF). 
Here we assembled a bespoke set of such problematic sites 
using the information in the Viridian sequences. 
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021, 
without any site masking. 
Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL). 
Only six of these sites overlapped with the “Problematic Sites” 
downloaded from the UCSC Genome Browser website (URL), 
which did not have a high number of mutations in the preliminary ARG. 
To infer the final ARG, we masked the problematic sites 
in our custom set (https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml).

@@

\subsubsection{Counting supporting loci for recombinant parents}
When counting such loci, we considered nearly adjacent sites (separated by at most three bases) as single loci, 
to avoid over-counting support from correlated loci, which would lead to artificial signals of recombination. 
Also, when counting supporting loci for a focal parent, 
we penalized loci where the character(s) in the recombinant matches that in the non-focal parent but not in the focal parent.

\subsubsection*{Calculating Pango lineage distance using pangonet}
We define a “pseudo-phylogenetic” distance to quantify the divergence between two Pango lineages. 
The idea is to count the number of edges separating two given Pango lineages in a “pseudo-phylogeny” 
which relates all the designated Pango lineages. 
To construct this pseudo-phylogeny, we used \textit{pangonet}, 
which mines the information on the Pango lineages in the “alias_key.json” and “lineage_notes.txt” files 
from the Pango designation issues (https://github.com/cov-lineages/pango-designation). 
If two Pango lineages are siblings in the pseudo-phylogeny, 
the distance is defined as the sum of the number of edges from each lineage to their MRCA; 
if they are non-siblings (i.e. one Pango is a descendant of the other), 
the distance is defined as the number of edges separating them; 
if they are identical, the distance is defined as zero.

@@

\subsubsection{Estimating interval lengths}
Recombination intervals are imprecise 
when the recombining parents are identical in sequence at sites spanning the breakpoint. 
Here we estimate the interval lengths expected around a breakpoint, 
given a genome-wide mutation rate $\mu$, 
total time between two samples ($T$), and 
sufficient support to detect the recombination event. 
If a breakpoint occurs a fraction $P$ along a genome of length $L$ and 
$k$ mutations have occurred to the left of the breakpoint, 
then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left. 
If we require $n$ mutations on both sides of the breakpoint, 
we can calculate the expected value of the interval to the left by summing across a Poisson-distributed number of mutations, 
conditioned on having at least $n$ mutations:

@@

\subsubsection{Analyzing global case counts}
To explore the timing of emergence of recombinants, 
case count data were downloaded from 
the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) 
at Johns Hopkins University \cite{Dong2020} and binned by week from January, 2020 to March, 2023. 
The proportion of each lineage $i$ in each week ($p_i$) was then inferred 
from the Scorpio designation at each node within the ARG for that week. 
Genetic variation was measured as the “expected heterozygosity” across lineages, 
which is the chance that two lineages chosen at random 
from that week had different Scorpio designations ($\sum_{i} \sum_{j} 2 p_i p_j$). 
The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept, 
because there may be unreported cases even without any reported cases. 
The slopes were highly significant, and 
the p-values were confirmed to be robust by randomly permuting the predictor variable 10,000 times. 
ANOVAs were conducted to determine which single predictor ($n$, $H$, $n H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week. 
These statistical tests and graphics were conducted in Mathematica 14.2 (Wolfram 2024) and available at GITHUB LINK.

>>>>

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model \cite{Li2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

@@

To illustrate the power of these libraries, we provide some examples
in which we compare the Usher and sc2ts intersections in tskit format described earlier.
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences
represents 31.34 GB of nucleotide calls.
Of the 80,574,518 missing data calls (Ns) in the alignments,
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code).
UShER imputed only 4 calls from this set incorrectly,
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 GB
(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format).
Performing this analysis using matUtils would require exporting the full dataset to VCF,
and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments.
In contrast with matUtils and most phylogenetic software,
which provide a command line interface and suite of predefined operations that typically output text files,
tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}.
As an illustration of the open-ended and efficient nature of the analyses this enables,
we computed the number of polytomies (nodes that have $> 2$ children)
and the maximum number of children per node in the Usher and sc2ts outputs.
In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
and compute the maximum number of children (5918).
In the sc2ts ARG, it took 385 ms and 6 lines of code to find the maximum number of polytomies (137,001) and
maximum number of children (7708) across all 348 local trees.

@@

Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
but has been understudied in part due to a lack of suitable data models, methods, and tools \cite{Neches2020}.
While the methods and tools developed here are tailored for SARS-CoV-2,
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is
that they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops.
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
making them highly adaptable for a broad range of applications.
Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers,
enabling and accelerating studies of SARS-CoV-2 for years to come.

@@

% \newpage

% \section*{MAIN FIGURE TITLES AND LEGENDS}

% %%%  At final submission, figure files MUST be
% %%%  provided separately as high-resolution image
% %%%  files. All of the panels for a figure should
% %%%  be in the same file. Figures should have
% %%%  clear labels/file names (Figure 1, Figure 2,
% %%%  etc.).

% %%%  Figure titles and legends should be placed
% %%%  at the end of the main text. You do not
% %%%  need to place the figures, nor their titles
% %%%  and legends, within the main text. While
% %%%  typesetting your article, our team will
% %%%  place each figure in the best location
% %%%  based on the final layout and on your
% %%%  figure citations, e.g., (Figures 1A and 1B).

% %%%  Please review the figure guidelines before
% %%%  submitting your final materials:
% %%%  https://www.cell.com/figureguidelines.

% \noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}

% \subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}

% The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
% \newline
% (A) Or each panel or group of panels can be described separately, as shown here and below.
% \newline
% (B) Graph of X, Y, and Z.
% \newline
% (C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
% \newline
% Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}. Example figure created by Cassie Comeau, Cell Press.

% \newpage

% \section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}

% %%%  Whenever possible, we encourage you to submit
% %%%  your main-text tables as Microsoft Word documents,
% %%%  using Word's table function. This will ensure
% %%%  the best results during conversion. Tables
% %%%  should be numbered Table 1, Table 2, etc. and
% %%%  should not include subpanels (do not use Table 1A,
% %%%  1B, etc.). Give each table a brief descriptive
% %%%  title. Table legends are optional but encouraged.
% %%%  Footnotes (superscript lowercase letters) should
% %%%  be used where necessary to indicate some feature
% %%%  of the data; please do not use bold, italic,
% %%%  colored text, or shading for this purpose. Use
% %%%  separate cells, not line breaks or spaces, for
% %%%  all discrete data elements. Small embedded
% %%%  graphics with color are OK.

% %%%  Like figures, all tables must be cited within
% %%%  the main text, and our typesetting team will
% %%%  place the tables within the typeset paper at
% %%%  the appropriate locations.

% \subsection*{Table 1. A table with clear organization of data}

% \begin{tabular}{|l | l | l | l|}
%  \hline
%  \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex]
%  \hline
%  Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex]
%  \hline
%  Row B & 7 & 78 & 5,415 \\ [1ex]
%  \hline
%  Row C & 545 & 778 & 7507 \\ [1ex]
%  \hline
%  Row D & 545 & 18,744 & 7,560 \\ [1ex]
%  \hline
%  Row E & 88 & 788 & 6,344 \\ [1ex]
%  \hline
% \end{tabular}

% \bigskip

% The table legend (optional) follows the table itself.
% The legend should be used to provide additional info that relates to the table as a whole.

% \textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table,
% such as this footnote to the first row (row A). Do not use footnotes in the table title.

% \textsuperscript{b}More footnotes

@@

\subsubsection*{Li and Stephens haplotype-copying model}

@@

In our early attempts to build ARGs, we explored using several values of $k$.
We ran sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020 \cite{Zhan2023}(REF).
We would expect that a good value for $k$ should lead to parsimonious ARGs
that capture plausible recombination events without incurring too many mutations, particularly reversions.
A value of $k$ too low could result in many artifactual recombination events,
whereas a value of $k$ too high could miss many genuine recombination events.
We found that ARGs built with $k$ set to 3 contained many dubious recombination events,
but ARGs built with $k$ set to 5 failed
to include the well supported recombination events documented in a previous study \cite{Jackson2021}.
Hence, we decided to use $k$ set to 4 to infer the final ARG described in this study.

@@

Retrospective sample grouping successfully captured most saltation lineages,
but integrating Delta and Omicron into the ARG required further manual intervention.
To ensure that Delta and Omicron are added, we introduce a third mechanism, “unconditional seeding”.
The idea is to choose suitable samples belonging to each VOC,
and attach them to the ARG unconditionally (i.e. without filtering using HMM cost).
We sought “seed” samples for Delta, Omicron BA.1, and BA.2.
For Delta, we used two sets of seeds:
(1) 8 samples assigned to B.1.617.2, and
(2) 9 additional samples assigned to its sibling lineage B.1.617.1.
Specifically, these samples are the earliest samples present in the Viridian v04 dataset
that were used for the designation of the VOCs.
For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively.
We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
as it could be well explained as as a non-recombinant given the ARG available on its collection date.
For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
which is the earliest BA.2 sample present in the Viridian v04 dataset.
It is dated later than SRR17041376 (the BA.1 seed),
allowing time for the BA.1 lineage to accumulate enough sampled descendants
and to get established in the ARG.
For the full list of seeds,
see
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}

@@

Sites which are enriched for systematic sequencing errors or frequently experience homoplasy
can cause artifactual signals of recombination (REF).
The bioinformatics community has identified and curated a set of problematic sites
that are recommended to be excluded before phylogenetic inference (REF).
Here we assembled a bespoke set of such problematic sites
using the information in the Viridian sequences.
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
Only six of these sites overlapped with the “Problematic Sites”
downloaded from the UCSC Genome Browser website (URL),
which did not have a high number of mutations in the preliminary ARG.
To infer the final ARG, we masked the problematic sites
in our custom set
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.

@@

\subsubsection*{Counting supporting loci for recombinant parents}
When counting such loci, we considered nearly adjacent sites (separated by at most three bases) as single loci,
to avoid over-counting support from correlated loci, which would lead to artificial signals of recombination.
Also, when counting supporting loci for a focal parent,
we penalized loci where the character(s) in the recombinant matches that in the non-focal parent but not in the focal parent.

\subsubsection*{Calculating Pango lineage distance using pangonet}
We define a “pseudo-phylogenetic” distance to quantify the divergence between two Pango lineages.
The idea is to count the number of edges separating two given Pango lineages in a “pseudo-phylogeny”
which relates all the designated Pango lineages.
To construct this pseudo-phylogeny, we used \textit{pangonet},
which mines the information on the Pango lineages in the ``alias\_key.json''
and ``lineage\_notes.txt'' files
from the Pango designation issues (https://github.com/cov-lineages/pango-designation).
If two Pango lineages are siblings in the pseudo-phylogeny,
the distance is defined as the sum of the number of edges from each lineage to their MRCA;
if they are non-siblings (i.e. one Pango is a descendant of the other),
the distance is defined as the number of edges separating them;
if they are identical, the distance is defined as zero.

@@

\subsubsection*{Estimating interval lengths}
Recombination intervals are imprecise
when the recombining parents are identical in sequence at sites spanning the breakpoint.
Here we estimate the interval lengths expected around a breakpoint,
given a genome-wide mutation rate $\mu$,
total time between two samples ($T$), and
sufficient support to detect the recombination event.
If a breakpoint occurs a fraction $P$ along a genome of length $L$ and
$k$ mutations have occurred to the left of the breakpoint,
then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left.
If we require $n$ mutations on both sides of the breakpoint,
we can calculate the expected value of the interval to the left by summing across a Poisson-distributed number of mutations,
conditioned on having at least $n$ mutations:

@@

\subsubsection*{Analyzing global case counts}
To explore the timing of emergence of recombinants,
case count data were downloaded from
the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE)
at Johns Hopkins University \cite{Dong2020} and binned by week from January, 2020 to March, 2023.
The proportion of each lineage $i$ in each week ($p_i$) was then inferred
from the Scorpio designation at each node within the ARG for that week.
Genetic variation was measured as the “expected heterozygosity” across lineages,
which is the chance that two lineages chosen at random
from that week had different Scorpio designations ($\sum_{i} \sum_{j} 2 p_i p_j$).
The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept,
because there may be unreported cases even without any reported cases.
The slopes were highly significant, and
the p-values were confirmed to be robust by randomly permuting the predictor variable 10,000 times.
ANOVAs were conducted to determine which single predictor ($n$, $H$, $n H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
These statistical tests and graphics were conducted in Mathematica 14.2 (Wolfram 2024) and available at GITHUB LINK.
